Making Sequential Programs Deterministic

### 2019-03-25 design.tex
poly resolution rules
small fix
fixes

Recall that the determinism checker does not permit \<@OrderNonDet> on a type that isn't a collection, iterator, or an array.
As a consequence, resolution of polymorphic type annotations at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{@PolyDet int size(@PolyDet List<T> list)}. Invoking this method
as \codeid{size(@OrderNonDet List<@Det String>)} will result in \<@PolyDet> getting resolved to \<@OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<@PolyDet> annotation, \<@PolyDet("up")>
and \<@PolyDet("down")>. The semantics of these variants are as follows:
\begin{itemize}
    \item \<@PolyDet("up")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@NonDet>.
    \item \<@PolyDet("down")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@Det>.
\end{itemize} 
The modifiers "up" and "down" are indicative of the subtyping relationship of \<@OrderNonDet> with \<@NonDet> and \<@Det> respectively
in the determinism type hierarchy.
Annotating \<size()> as \codeid{@PolyDet("down") int size(@PolyDet List<T> list)} will only allow the following type resolutions:
\begin{itemize}
    \item \codeid{@Det int size(@Det List<T> list)}
    \item \codeid{@Det int size(@OrderNonDet List<T> list)}
    \item \codeid{@NonDet int size(@NonDet List<T> list)}
\end{itemize}

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
The checker framework performs automatic type inference for local types within method bodies.
It also performs dataflow analysis and refines the types of these locals, thereby reducing the annotation burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       //  No error here.
}
\end{verbatim}
The checker framework performs dataflow analysis within method bodies and refines the types of local variables.
In the example above, even though the local variable \codeid{y} was declared to be \codeid{@NonDet}, it gets
type refined to \codeid{@Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{@Det} variable \codeid{z} is sound. In the following example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
the checker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

The determinism checker has additional type refinement rules for the sorting methods in collections. 
The checker type refines a receiver of type \codeid{@OrderNonDet List} to a \codeid{@Det List} if all the
type argument of this \codeid{List} have the type annotation \codeid{@Det}. Similar rules apply for \codeid{Arrays.sort()} and
\codeid{Collections.sort()}.
Under this if \codeid{sort()} is called on a receiver of type \codeid{@OrderNonDet List<@Det Set<@Det String>>}, it will
be type refined to \codeid{@Det List<@Det Set<@Det String>>}. However a receiver of type \codeid{@OrderNonDet List<@OrderNonDet Set<@Det String>>} will not be type refined.

>>>>

Recall that the determinism checker does not permit \<@OrderNonDet> on a type that isn't a collection, iterator, or an array.
As a consequence, resolution of polymorphic type annotations at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{@PolyDet int size(@PolyDet List<T> list)}. Invoking this method
as \codeid{size(@OrderNonDet List<@Det String>)} will result in \<@PolyDet> getting resolved to \<@OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<@PolyDet> annotation, \<@PolyDet("up")>
and \<@PolyDet("down")>.

Valid locations for \codeid{@PolyDet("up")} and  \codeid{@PolyDet("down")}:

$\infer{\vdash : \kappa \  \tau}{isParam(\tau) / isReturn(\tau) & \kappa : @PolyDet("up") / @PolyDet("down")}$

Resolution of \<@PolyDet("up")>:\todo{typesetting}

$\infer{\Gamma \vdash x : @NonDet \ \tau_a}{\Gamma \vdash x : \kappa_a \ \tau_a &  \kappa_a : @OrderNonDet/@NonDet & \kappa_p : @PolyDet("up") & \kappa_p \ \tau_p : declaration \ type & \kappa_a \ \tau_a : resolved \ type}$

Resolution of \<@PolyDet("down")>:

$\infer{\Gamma \vdash x : @Det \  \tau_a}{\Gamma \vdash x :  \kappa_a \tau_a & \kappa_a : @OrderNonDet/@Det & \kappa_p : @PolyDet("down") & \kappa_p \ \tau_p : declaration \ type & \kappa_a \ \tau_a : resolved \ type}$

The semantics of these variants are as follows:
\begin{itemize}
    \item \<@PolyDet("up")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@NonDet>.
    \item \<@PolyDet("down")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@Det>.
\end{itemize} 
The modifiers "up" and "down" are indicative of the subtyping relationship of \<@OrderNonDet> with \<@NonDet> and \<@Det> respectively
in the determinism type hierarchy.
Annotating \<size()> as \codeid{@PolyDet("down") int size(@PolyDet List<T> list)} will only allow the following type resolutions:
\begin{itemize}
    \item \codeid{@Det int size(@Det List<T> list)}
    \item \codeid{@Det int size(@OrderNonDet List<T> list)}
    \item \codeid{@NonDet int size(@NonDet List<T> list)}
\end{itemize}

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
The checker framework performs automatic type inference for local types within method bodies.
It also performs dataflow analysis and refines the types of these locals, thereby reducing the annotation burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       //  No error here.
}
\end{verbatim}
In the example above, even though the local variable \codeid{y} was declared to be \codeid{@NonDet}, it gets
type refined to \codeid{@Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{@Det} variable \codeid{z} is sound. In the following example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
the checker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

The determinism checker has additional type refinement rule for the sorting methods in collections.

$\infer{}{}$

The checker type refines a receiver of type \codeid{@OrderNonDet List} to a \codeid{@Det List} if all the
type argument of this \codeid{List} have the type annotation \codeid{@Det}. Similar rules apply for \codeid{Arrays.sort()} and
\codeid{Collections.sort()}.
Under this if \codeid{sort()} is called on a receiver of type \codeid{@OrderNonDet List<@Det Set<@Det String>>}, it will
be type refined to \codeid{@Det List<@Det Set<@Det String>>}. However a receiver of type \codeid{@OrderNonDet List<@OrderNonDet Set<@Det String>>} will not be type refined.
